### 2019-06-11 report/chapter04.Rnw
Methods update

\chapter{Methods}

\section{Basic notation}
The notation used here will be used throughout the chapter and exceptions will be noted. Let $i$ be the number of a study of a meta analysis with $n$ being the total number of studies. $y_{i}$ is then the effect size estimate (usually log odds ratio or mean difference) and $v_{i}$ the variance of the estimate of study $i$. $w_{i}$ is used for weights which are defined when necessary, and $\Delta$ usually denotes the summarized or pooled effect estimate of the meta-analysis, and $\eta$ the variance thereof.

\vspace{0mm}
In the case of binary outcomes, let $e_{t}$ be the number of events and $n_{t}$ be the total number of patients in the treatment arm and $n_{c}$ and $e_{c}$ analogously for the control arm in a two-armed study $i$.

\section{Heterogeneity}
In addition to sampling error, there can be additional, ``real'' variation between estimates of different studies, indicating real differences between the studies. This is called between study variation in contrast to within study variation (noise).

\vspace{0mm}
The $Q$ statistic is a weighted sum of squares that quantifies the deviation from the weighted mean of study effect estimates. Let $w_i$ be the inverse of the variance and $\Delta$ be a summarized effect estimate of your choice as for example a variance-weighted mean \ref{weighted.mean}. Then $Q$ can be calculated as in \ref{Q.heterogeneity}

\begin{align}
\Delta &= \frac{\sum_{i = 1}^n w_{i}y_{i}}{\sum_{i = 1}^n w_{i}} \label{weighted.mean} \\
Q &= \sum_{i = 1}^n w_{i}(y_{i} - \Delta)^2 \label{Q.heterogeneity}
\end{align}

Because $Q$ is a standardized measure, it does not depend on the effect size, but only on the study number $n$. Under the assumption of equal effect sizes of all studies, the expected value of $Q$ is $n-1$, so the excess dispersion is just $Q - n + 1$.To test the assumption of equal effect sizes one uses that $Q$ follows a central Chi-squared distribution with $n -1$ degrees of freedom under the null hypothesis of equal effect sizes. $1 - F(Q)$ will provide the $p$-value for the significance test with $F$ being the cumulative distribution function of the Chi-squared distribution with the corresponding degrees of freedom.

\vspace{0mm}
Since $Q$ is a standardized metric, it gives no impression of the real dispersion of the effect sizes. For this purpose, $\tau^2$, the variance of true effects, can be calculated. $\tau^2$ is on the same scale as the effect size and reflects the absolute amount of dispersion. In practice, $\tau^2$ can be smaller than zero, then it is set to zero.

\begin{align}
C &= \sum_{i = 1}^n w_{i} - \frac{\sum_{i = 1}^n w_{i}^2}{\sum_{i = 1}^n w_{i}} \label{C.definition} \\
\tau^2 &= \max(0, \frac{Q - d}{C}) \label{Tau.definition}
\end{align}

The estimation method for $\tau^2$ is known as DerSimonian and Laird method, but others, such as restricted maximum likelihood can be used. Note that their estimate can differ substantially and consequently also the estimate of the pooled effect size estimate.

\vspace{0mm}
To estimate the proportion of real variance between effect estimates of the observed variance, the $I^2$ can be used. The calculation is given in \ref{I2.proportion}

@@

There are ways to compute confidence intervals for $I^2$ and $\tau^2$ that are not shown (see \cite[122]{metaanalysis}).

\section{Meta Analysis}
There are numerous methods to pool the estimates of multiple studies into one estimate, and two will be introduced here; fixed and random effects meta-analysis.
First the fixed effect meta-analysis will be explained. Note that both methods can be used for continuous or dichotomous outcomes. For more details about the methods, see chapter 11 and 12 in \citet{metaanalysis}

\vspace{0mm}
Let $w_i = 1/v_{i}$ be the inverse of the variance of the estimate from study $i$. The pooled estimate $\Delta_{f}$ of the fixed effects model is then the weighted average with the weights given by the inverses of the variances, $w_{i}$, given in \ref{weighted.mean}. The variance $\eta_{f}$ is the reciprocal of the sum of the weights as shown in \ref{reciproce.variance}.

\begin{align}
\Delta_{f} &= \frac{\sum_{i = 1}^n w_{i}y_{i}}{\sum_{i = 1}^n w_{i}} \label{weighted.mean} \\
\eta_{f} &= \frac{1}{\sum_{i = 1}^n w_{i}} \label{reciproce.variance}
\end{align}

The computation of random effects meta-analysis is more complicated. Random effects meta-analysis will give smaller studies with larger variance more weight in the pooled estimate. Shortly, the idea is that the estimates are allowed to vary randomly around the true estimate $\Delta$, and additionally, the estimates are subject to noise or sampling error themselves.

\vspace{0mm}
The variance of a study estimate $y_{i}$ of study $i$, $v_{i}^\star$ is defined as in \ref{ranef.study.variance}, with $w_{i}^\star$ being the inverse of $v_{i}^\star$. It is used to calculate a new weighted mean to obtain a pooled estimate $\Delta_{r}$ as in \ref{ranef.weighted.mean}. The variance of $\Delta_{r}$, $\eta_{r}$ is then the sum of the reciprocal variances (\ref{ranef.reciproce.variance}).

\begin{align}
v_{i}^\star &= v_{i} + \tau^2 \label{ranef.study.variance} \\
\Delta_{r} &= \frac{\sum_{i = 1}^n w_{i}^\star y_{i}}{\sum_{i = 1}^n w_{i}^\star} \label{ranef.weighted.mean} \\
\eta_{r} &= \frac{1}{\sum_{i = 1}^n w_{i}^\star} \label{ranef.reciproce.variance}
\end{align}

A p-value under the Null-hypothesis of $\Delta = 0$ can be obtained by calculating the $Z$-value (\ref{Zvalue}) and using the distribution function of a standard normal as shown in (\ref{p.value.calculation}), $\Phi$ being the distribution function of a standard normal distribution.

\begin{align}
Z &= \frac{\Delta}{\sqrt{\nu}} \label{Zvalue} \\
p &= 2(1 - \Phi(|{Z}|) \label{p.value.calculation}
\end{align}

\section{Small Study Effects Tests}
One crucial assumption in meta analysis is that the availability and publication of studies does not depend on their effect and the variance of the effect. If this is not given, one often speaks of publication bias. In fact, there can also be other reasons for this (see discussion section). A more appropriate term for the phenomenon is small study effect. If small study effects are present in a meta-analysis, the classical approaches to merge single study results in to an overall intervention effect fails to provide an appropriate estimate of the treatment effect.

\subsection{Continuous Outcome Tests}
% Essentially, there are two kinds of tests for reporting bias, non-parametrical or rank-based tests or regression based tests. 
% First, tests for continuous outcome studies are described, and special modifications of those for binary outcomes will be introduced later.

\subsubsection{Begg and Mazumdar: Rank Correlation Test}
\citet{begg.ties} proposed a rank based test to test the null hypothesis of no correlation between effect size and variance.
A standardized effect size $y_{i}^\star$ can be computed as in \ref{begg.stand.eff}. $v_{i}^\star$ is the variance of $y_{i} - \Delta_{f}$ as defined in \ref{begg.stand.var}. $\Delta_{f}$ is the pooled estimate for fixed effect estimate defined in \ref{weighted.mean}.

\begin{align}
y_{i}^\star &= (y_{i} - \Delta_{f})/v_{i}^\star \label{begg.stand.eff} \\
v_{i}^\star &= v_{i} - 1/\sum_{i = 1}^n v_{i}^-1 \label{begg.stand.var} 
\end{align}

A rank correlation test based on Kendall's tau is then used. The pairs $(y_{i}^\star, v_{i}^\star)$ that are ranked in the same order are enumerated. Let $u$ be the number of pairs ranked in the same order, and $l$ the number of pairs ranked in the opposite order (e.g. larger standardized effect size and smaller variance). Then the normalized test statistic $Z$ is given in \ref{begg.teststat}.

\begin{align}
Z &= (u - l)/\sqrt{n(n-1)(2n + 5)/18} \label{begg.teststat}
\end{align}

The changes in the case of ties are negligible \cite[410]{begg.ties}.

\subsubsection{Egger's Test: Linear Regression Test}
Alternatively, one can use Eggers test \citep{Egger} that is based on linear regression. Let $y_{i}^\star = y_{i}/\sqrt{v_{i}}$ and $x_{i} = 1/\sqrt{v_{i}}$. 
Using $y_{i}^\star$ as dependent, and  $x_{i}$ as explanatory variable in linear regression, one obtains an intercept $\beta_{0}$ and a slope.

\vspace{0mm}
If $\beta_{0} \ne 0$, the null hypothesis of no small study effect may be contested, using that $\beta_{0} \sim t_{n-1}$, $n-1$ being the degrees of freedom of the $t$-distribution. The p-value for $\beta_{0} = 0$ (no reporting bias) is then given by \ref{egger.pvalue}.

\begin{align}
p &= 2*(1 - t_{n-1}(\beta_{0}/se(\beta_{0}))) \label{egger.pvalue}
\end{align}

\subsubsection{Thompson and Sharp's Test: Weighted Linear Regression Random Effects Test} \label{Thompson}
A method proposed in \citet{thompson.sharp} allows for between study heterogeneity. Let $\tau^2$ be equal to \ref{Tau.definition}. The effect size estimates are then assumed to be distributed as in \ref{t.sharp.regression}.

\begin{align}
y_{i} \sim N(\beta_{0} + \beta_{1}x_{i}, v_{i} + \tau^2) \label{t.sharp.regression}
\end{align}

Then, a weighted regression is carried out with weights $1/v_{i}^\star$ based on the inverse of the variance as in \ref{ranef.study.variance}. Analogous to Egger's test, $\beta_{0}$ is then tested with respect to the null hypothesis $\beta_{0} = 0$.

@@

The issue with dichotomous outcomes is that effect size and variance of effect size are not independent. Consequently, the tests above will tend to reject the null-hypothesis too often, i.e. that they are not conservative enough. A number of solutions to this problem are existing in the literature.

\subsubsection{Peters Test: Weighted Linear Regression Test}
A modification of the weighted linear regression test that takes into account effect size and variance interdependence for dichotomous outcomes is proposed in \citet{Peters}.

\vspace{0mm}
Let $y_i$ be the log-odds ratio estimate \ref{log.odds.ratio} and $v_{i}$ its variance \ref{variance.log.odds.ratio}

\begin{align}
y_{i} &= \log(e_{t}*(n_{c} - e_{c})/e_{c}*(n_{t} - e_{t})) \label{log.odds.ratio}\\
v_{i} &= 1/(e_{t}+(n_{t} - e_{t}) + 1/(e_{c}+(n_{c} - e_{c}) \label{variance.log.odds.ratio}
\end{align}

and $x_{i}$ be the total sample size $n_{t} + n_{c}$. Instead of taking the variance as explanatory or independent variable in regression as in Egger's Test, the inverse of the total sample size $x_{i}$ is used, and the variance $v_{i}$ is used as a weight. The subsequent test procedure is then identical to Egger's test.

\vspace{0mm}
Peters test is a a small modification of Macaskill's test where the explanatory variable is the sample size instead of its inverse.

\subsubsection{Harbord's Test: Score based Test}
A rank based alternative to Peters test for binary outcomes is the Harbord's test \citep{Harbord}. The score $r_{i}$ (the first derivative of the log-likelihood of a proportion with treatment effect equal 0) and its variance $v_{i}$ can be computed as shown in \ref{harbord.score} and \ref{harbord.variance}.

>>>>

\chapter{Methods}
\section{Introduction and Notation}
The analysis has already been said to be restricted on clinical or health care interventions. The interventions are restricted to two participant arms of which some measure of sanity or worsening of health is measured and compared. The difference between these two is referred to as the effect of the treatment. Where it is not particularly mentioned, the term treatment effect refers to any effect measure such as log risk ratio, log hazard ratio, Cohen's $d$ or Hedges $g$, fisher's z score or pearson correlation. \\
Let us consider a meta-analysis with $n$ study treatment effects ($n > 1$, but typically small). A study is indexed by $i$, and it's treatment effect by  $\theta_i$. The observed treatment effect is $\hat{\theta}_i$. The pooled treatment effect of a meta-analysis will be denoted as $\theta_M$, and consequently, the observed pooled treatment effect as $\hat{\theta}_M$. Furthermore, each treatment effect is typically measured with some standard error $s_i$ and an estimate of $s_i$ is given by $\hat{s}_i$. \\
For continuous outcomes, let $m_t$ be the mean of the treatment group, $m_c$ the mean of the control group, and equivalentyl $sd_t$ and $sd_c$ the corresponding standard deviations. $n_t$ and $n_c$ are the total number of participants in the groups.
In the case of binary outcomes, let $e_t$ be the count of events in the treatment arm $e_c$ the count of events in the control group. The observed counts in a study $i$ are referred to as $e_{t,i}$ and analogously $e_{c,i}$.

\section{Effect Measures and p-values}
\subsection{Continuous Outcomes}
For given $(m_t, m_c), (sd_t, sd_c)$ and $(n_t, n_c)$, one can compute mean difference as well as a standardized mean difference (here: Hedges $g$) and  a standard error thereof. Note that the definition of Hedges $g$ and its standard error $s$ varies amongst the literature, the following applies for this report:

\begin{align}
s &= \sqrt{\frac{(n_t - 1)sd_t^2 + (n_c - 1)sd_c^2}{n_t + n_c - 2}} & g &= \frac{m_t - m_c}{s} \label{eq:hedges.g}
\end{align}

The mean difference $\theta$ and its standard error $s$ can similarly be obtained by

\begin{align}
\theta &= m_t - m_c & s &= \sqrt{sd_t^2/n_t + sd_c^2/n_t}
\end{align}

Both estimators take into account that the two groups might have unequal variances. A p-value to test the null hypothesis that the mean between group is equal is commonly obtained with the students $t$ test. The $t$ statistic is obtained, using $s$ and $g$ from \ref{eq:hedges.g}, by

\begin{align}
t &= g/(s\sqrt{(1/n_t)+(1/n_c)}) \nonumber
\end{align}
and the $p$-value can be obtained with the cumulative student's $t$-distibution $F$ with $n_t + n_c - 2$ degrees of freedom:

\begin{align}
p &= 2(1-F(|t|)) \nonumber
\end{align}

The t-test is known to be less reliable if combined sample size is small ($n < 30$), see for example \citet{t.test}.

\subsection{Binary Outcomes}
Two commonly used effect measures for binary outcome data are risk ratios and odds ratios between treatment and control group. 
The methods presented here can also be found, for example in \citet{Intro.meta}, pg. 34.
Let $\theta$ be the logarithm to base 10 of the odds ratio. $\hat{\theta}$ and its variance $s^2$ can be obtained by

\begin{align}
\hat{\theta} &= \log(\frac{e_t*(n_c - e_c)}{e_c*(n_t - e_t)}) \nonumber \\
\hat{s}^2 &= 1/e_t + 1/(n_t - e_t) + 1/e_c + 1/(n_c - e_c) \nonumber
\end{align}

The logarithm of the risk ratio $\theta$ and its variance $s^2$, can be estimated by

\begin{align}
\hat{\theta} &= \log(\frac{e_t/n_t}{e_c/n_c}) \label{eq:risk.ratio} \\
\hat{s^2} &= 1/e_t + 1/n_t + 1/e_c + 1/e_t \label{eq:risk.ratio.variance}
\end{align}

Using likelihood theory, one could show that the estimates are maximum likelihood estimates and that one can use the asymptotic normal distribution of the maximum likelihood estimator to calculate a $p$-value (\citet{held2014}, pg.98). \\
Thus, with $\Phi$ as the cumulative standard normal distribution, we get
\begin{align}
p &= 2*(1-\Phi(|\hat{\theta}/\hat{s}|) \nonumber
\end{align}
which summarizes the evidence against $\theta$ being zero (i.e. the true risk/odds ratio being 1). \\
Binary and continuous effect measures can be converted into each other. In section \ref{sec:transformation.effectsizes} at the end of this chapter, this is readily described.

\subsection{Survival Outcomes}
Time-to-event data with censoring present have to be analyzed by special means. One frequently used method to take into account right-censoring is the Cox proportional hazards regression model \citep{Cox}. Because the method itself is not applied in this thesis, but only the resulting estimates of the parameters are used, the reader is referred to the extensive literature covering this topic (e.g. \citet{Surv}).
The so-called hazard ratio estimated by cox p.h. regression is the ratio of the instanteneous risk of experiencing the event between the two groups. Because it is a maximum likelihood estimator, one can again use its wald test statistic to test for equal hazards. \\
Let $\hat{\theta}$ be an estimate of the log hazard ratio and $\hat{s}$ an estimate of the standard error of it. As before
\begin{align}
p &= 2*(1-\phi(|\hat{\theta}/\hat{s}|) \nonumber
\end{align}
will give a $p$-value for the evidence against the null hypothesis.

\section{Fixed and Random Effects Meta-Analysis} \label{sec:meta.analysis}
The fixed effects meta-analysis estimator of the pooled treatment effect is a mean of the single treatment effect estimators, weighted by their standard errors \cite{fixed.effects.rosenthal}. Let $w_i = 1/s_i^2$ be the weights, and $\theta_M$ be the pooled estimator and $s_M^2$ its variance. Then

\begin{align}
\theta_M &= \frac{\sum_{i = 1}^n w_i \theta_i}{\sum_{i = 1}^n w_i} &
s_M^2 &= \frac{1}{\sum_{i = 1}^n w_i} \label{eq:fixed.effects}
\end{align}

This estimator minimizes the variance, an estimate $\hat{\theta}_M$ can be obtained by plugging in the observed treatment effects and standard errors $\hat{\theta}_i, \hat{s}_i^2$. The underlying idea is that we assume $\hat{\theta}_i \sim N(\theta_M, s_i^2)$, $\theta_M$ being the true effect, such that all $\theta_i$ are disitributed equally. \\
The random effects model \citep{whitehead} assumes instead that 
\begin{align}
\theta_i \sim N(\mu_i, s_i^2) &&
\mu_i \sim N(\theta_M, \tau^2) \label{eq:random.effects} 
\end{align}

So marginally, we have $\theta_i$ being distributed around a common mean $\theta_M$ with additional variance $\tau^2$:

\begin{align}
\theta_i | \mu_i \sim N(\theta_M, s_i^2 + \tau^2) \label{eq:random.effects.marginal}
\end{align}

$\tau^2$ is often referred to as a population variance or between-study variance, whereas $s_i^2$ can be interpreted as sampling error. The pooled treatment effect and its variance is obtained by replacing the weights $w_i$ in equation \ref{eq:fixed.effects} with $w_i = 1/(s_i^2 + \tau^2)$. \\
The model is clearly superior to the fixed effects model whenever the standard errors of the treatment effects alone are unlikely to fully account for the entire variability between studies. Note that as $\tau^2$ increases, each $\theta_i$ will eventually get equal weights, irrespective of sampling error.\\
The estimation of $\tau^2$ has been subject to some debate in the statistical literature. Oftentimes, the method of moment estimator of \citet{tau.estimator} is used.
We use the measure of heterogeneity, $Q$, and divide by $C$ after having subtracted the degrees of freedom $n-1$:

\begin{align}
Q &= \sum_{i = 1}^n w_i(y_i - \theta_M)^2 & C &= \sum_{i = 1}^n w_i - \frac{\sum_{i = 1}^n w_i^2}{\sum_{i = 1}^n w_i}\label{eq:Q.heterogeneity} \\
\tau^2 &= \max(0, \frac{Q - (n-1)}{C}) \label{eq:Tau.definition}
\end{align}

Again, $w_i, \theta_i ...$ have to be replaced by their estimates in order to get an estimate $\hat{\tau}^2$. \\
The Paule-Madel estimator \citep{paulemandel} is considered to have most often better properties than the method of moments estimator \citep{tau.estimator.evaluation}.
Since we defined $w_i = 1/(s_i^2 + \tau^2)$, it is straightforward that

% \begin{align}
% w_i \textrm{Var}(\theta_i) &= 1 & \textrm{Var}(\sqrt{w_i}\theta_i) = 1 \nolabel
% \end{align}

For any $w_i$, the variance can be estimated and equated to its expected value:

\begin{align}
s^2(w_i\theta_i) &= \frac{\sum_{i = 1}^n w_i(\theta_i - \theta_M)^2}{n-1} & \frac{\sum_{i = 1}^n w_i(\theta_i - \theta_M)^2}{n-1} = 1 \label{eq:paulemandel}
\end{align}

$\theta_M$ can be estimated using equation \ref{eq:fixed.effects} and the new weights $w_i$, thus, the only problem is to estimate $\tau^2$. It can be obtained through an iterative process, using a newly defined function

\begin{align}
F(\tau^2) &= \sum_{i = 1}^n w_i(\theta_i - \theta_M)^2 - (n-1) \nonumber
\end{align}

In view of equation \ref{eq:paulemandel}, $\tau^2$ must be such that $F(\tau^2) = 0$. We start with a arbitrary $\tau^2$, iteratively add a term $\tau_0^2$ and update $\tau^2$ until $F(\tau^2 + \tau_0^2)$ is close to zero. Using a truncated taylor series expansion, one can obtain the partial derivative after $\tau^2$, which is a reasonable choice for $\tau_0^2$. Using $\tau^2 + \tau_0^2$ for $\hat{w}_i, \hat{theta}$, we can update $F(\tau^2)$ and check convergence to zero. \\
The estimation of $\tau^2$ is accompained by uncertainity. A common procedure is to test if it is there is significant heterogeneity between the studies (\citet{Intro.meta}, pg. 109). Compute $Q$, as given in \ref{eq:Q.heterogeneity}, based on one of the estimators of $\tau^2$. It is assumed that $Q$ follows a central Chi-squared distribution with $n -1$ degrees of freedom under the null hypothesis of equally distributed effect sizes. Thus, the expected value of $Q$ is $n-1$, and the excess dispersion is $Q - n + 1$. The $p$-value against the null hypothesis of equally distributed effect sizes is, using $F$ as the cumulative distribution function of the Chi-squared distribution with d.f. = $n-1$, $1 - F(Q)$. \\
An advantage of the $\tau^2$ is that it is directly linked to the variability in the data. Additionally, one can use the $I^2$ statistic to see what portion the between study variance has of the overall variance. It is computed as

@@

Large values for $\hat{I}^2$ are of great importance when analysing e.g. small study effects. \\
Importantly, all proposed methods above assume normality and proper estimation of standard errors. This assumptions are not met for very small sample sizes and very few event counts. An alternative in the latter case is the mantel-haenszel method for risk and odds ratios (see e.g. \citet{mantel.haenszel}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Small Study Effects Tests}
The tests that will be presented on the following pages are a common way to detect publication bias. Importantly, they are however not interpretable directly as evidence for publication bias, but this is left for the discussion chapter. The tests are also often referred to as funnel plot asymmetry tests, because of the popularity of a recent test that has been used frequently to test and adjust for publication bias, which goes under the name of trim-and-fill \citet{trimfill}. However, it is known for some time that the method has disadvantageous properties, therefore, it will not be discussed here, as well as the funnel plot (the radial plot will be used as an alternative).\\
A test for small study effects will test in some ways if the size of the estimated treatment effect of a study depends on some measure of its size. An association between study size and treatment effect size can be interpreted as an artifact of publication bias.

\subsection{Continuous Outcome Tests}
For a continuous outcomes that are normally distributed, the sample mean and variance are independent of each other (\citet{meta.w.R}, pg 120). Thus, the estimated standard errors of treatment effects can be used as a proxy for study size, as they should in principle not be tied to the effect size.

\subsubsection{Begg and Mazumdar: Rank Correlation Test}
\citet{begg.ties} proposed a rank based test to test the null hypothesis of no correlation between effect size and variance.
A standardized effect size $\theta_i^\star$ can be computed as in \ref{begg.stand.eff}. $s_i^{2\star}$ is the variance of $\theta_i - \theta_M$ as defined in \ref{begg.stand.var}, $\theta_M$ being the fixed effects pooled treatment effect (\ref{eq:fixed.effects}.

\begin{align}
\theta_i^\star &= (\theta_i - \theta_M)/s_i^{2\star} \label{begg.stand.eff}  \\
s_i^{2\star} &= s_i^2 - 1/\sum_{i = 1}^n\frac{1}{s_i^2}1 \label{begg.stand.var} 
\end{align}

A rank correlation test based on Kendall's tau is then used. First, the pairs are ordered after their ranks based on $s^{2\star}$. Then, for each $s^{2\star}$ rank, the corresponding ranks based on $\theta^{2\star}$ that are larger are counted and summed up to $u$. The number of ranks based on $\theta^{2\star}$ that are in contrary, smaller, are counted and summed up to $l$. Then the normalized test statistic $Z$ is given as

\begin{align}
Z &= (u - l)/\sqrt{n(n-1)(2n + 5)/18} \nonumber
\end{align}

Thus, large number of concordance between pairs will reflect in large $\hat{u}$ and small $\hat{l}$ and thus lead to a large $\hat{Z}$. The $p$-value is obtained using the standard normal distribution $\Phi$:
\begin{align}
p &= 2*(1-\Phi(|Z|)) \nonumber
\end{align}

The changes that have to be made int the case of ties are small and can be found in \cite[410]{begg.ties}.

\subsubsection{Egger's Test: Weighted Linear Regression Test}
First, the concept of simple linear regression is introduced. In short, the model assumes a dependent variable $y$ to be a linear function of another explanatory variable $x$:

\begin{align}
y &= \beta_0 + \beta_1 x + \epsilon, & \epsilon \sim N(0, \sigma^2) \label{eq:simple.regression}
\end{align}

$\epsilon$ is the residual noise term that becomes necessary when $n$ pairs $(x_i, y_i)$ are given and there is no exact solution. Then it is common to look for the solution that minimizes the squared residuals, the least-squares solution. Formally,

\begin{align}
\operatorname*{argmin}_{\beta_0, \beta_1}(\sum_{i = 1}^n y_i - \beta_0 - \beta_1 x_i) \label{eq:least.squares}
\end{align}

Let $\mathbf{X}$ be a matrix whith the explanatory variables $x$ and $\mathbf{y}$ a corresponding vector for all $y$:

\begin{equation*}
\mathbf{X} = 
\begin{bmatrix}
1 & x_{12} \\
1 & x_{22} \\
1 & x_{32} \\
\vdots & \vdots \\
1 & x_{n2} \\
\end{bmatrix} 
\qquad
\mathbf{y} = 
\begin{bmatrix}
y_1 \\
y_2 \\
y_3 \\
\vdots \\
y_n \\
\end{bmatrix}
\end{equation*}

Let $\mathbf{\beta} = (\beta_0, \beta_1)^\top$. It can be shown that

\begin{align}
\hat{\mathbf{\beta}} &= (\mathbf{X}^\top \mathbf{X})^{-1} \mathbf{X}^\top \mathbf{y} \label{eq:regression.parameters}
\end{align}

Is an estimator of $\mathbf{\beta}$ that mimizes the squared residuals. Let $\hat{\mathbf{y}} = \mathbf{X}\hat{\mathbf{\beta}}$ and $\hat{\mathbf{r}} = \hat{\mathbf{y}} - \mathbf{y}$. The variance estimates $\hat{\sigma}^2$ and $\hat{\mathbf{s}}_\beta^2$ are

\begin{align}
\hat{\sigma}^2 &= \frac{1}{n-2}\mathbf{r}^\top \hat{\mathbf{r}} & \hat{\mathbf{s}}_\beta^2 &= \hat{\sigma}^2 (\mathbf{X}^\top \mathbf{X})^{-1} \label{eq:regression.variances}
\end{align}

The estimate $\hat{\beta}_0$ is the intercept and $\hat{\beta}_1$ the slope of the regression line. Furthermore, in the simple linear regression setting, $\hat{\beta}_0$ can also be obtained by:

\begin{align}
% \hat{\beta_1} &= \frac{\sum_{i = 1}^n (x_i - \bar{x})(y_i - \bar{y})}{\sum_{i = 1}^n (x_i - \bar{x})^2} &
\hat{\beta}_0 &= \bar{y} - \hat{\beta}_1 \bar{x} \nonumber
\end{align}

$\bar{x}, \bar{y}$ denoting the sample means of the corresponding values $x_1, .., x_n$ and $y_1, ..., y_n$. Thus, $\hat{\beta}_0$ is also called global mean.
To test whether there is evidence for the intercept $\beta_0$ to be unequal to some value $\beta_{H0}$, a $t$-test can be used.

\begin{align}
p &= 2(1-F(|(\beta_0 - \beta_{H0})/s_{\beta_0}|)) \nonumber
\end{align}

where $F$ is the cumulative $t$ distribution with $n-2$ degrees of freedom. $p$ will give the evidence against the null hypothesis $\beta_0 = \beta_{H0}$. \\
The concept is extendable to weigthed linear regression. Weigthed linear regression may be used if the residuals $\mathbf{r}$ have unequal variances, which is equivalent to ascribe different precision to the observed $y$. The least squares equation \ref{eq:least.squares} is extended to

\begin{align}
\operatorname*{argmin}_{\beta_0, \beta_1}(\sum_{i = 1}^n w_i(y_i - \beta_0 - \beta_1 x_i)) \nonumber
\end{align}

with positive weigths $w_i$ that penalize large squared residuals for some $i$ more if $w_i$ is larger compared to other $w_i$. \\
Let $\mathbf{W}$ be a $n \times n$ matrix with $\mathbf{W}_{ii} = w_i$, the weigths on the diagonal and zeros on the off-diagonals. The estimates in \ref{eq:regression.parameters} and \ref{eq:regression.variances} can stil be used if $\mathbf{X}$ is exchanged with $\mathbf{X}^\star =  \mathbf{W} \mathbf{X}$ and $\mathbf{y}$ with $\mathbf{y}^\star = \mathbf{W} \mathbf{y}$. \label{weighted.regression} \\
Now it is shown how linear regression can be used to test dependency of effect sizes on study sizes. The simplest application was introduced by \citet{Egger}.
Let $\theta/s$ be the dependent variable $y$ and $1/s$ the explanatory variable $x$. If plotted, this corresponds to a radial or galbraith plot \citet{galbraith}. The linear regression equation as introduced before in \ref{eq:simple.regression} can be written in two ways:

\begin{align}
\theta/s &= \beta_0 + \beta_1/s + \epsilon, & \epsilon \sim N(0, \sigma^2) \label{eq:radial.plot} 
\end{align}

\ref{eq:radial.plot} is often provided due to the correspondance to the radial plot. However, it is equivalent to

\begin{align}
\theta &= \beta_0 + \beta_1 s + \epsilon, & \epsilon \sim N(0, w^{-1}\sigma^2) \label{eq:egger.plot}
\end{align}

with weigths $w = 1/s^2$. Thus testing $\beta_0$ of \ref{eq:radial.plot} or $\beta_1$ is equivalent. The corresponding $p$-value is then used as evidence for a small study effect. Plugging in  $\theta_i/s_i, 1/s_i$ as $y_i, x_i$ into equation \ref{eq:regression.parameters} and \ref{eq:regression.variances} will give the estimates for $\hat{\beta}_0, \hat{\beta}_1,\hat{s}_{\beta_0}$ and $\hat{s}_{\beta_1}$.

% \begin{align}
% \hat{\beta_1} &= \frac{\sum_{i = 1}^n (\varphi_i - \bar{\varphi})(\theta^\star_i - \bar{\theta}^\star)}{\sum_{i = 1}^n (\varphi_i - \bar{\varphi})^2} &
% \hat{\beta}_0 &= \bar{\theta}^\star - \hat{\beta}_1 \bar{\varphi} \nonumber
% \end{align}

\subsubsection{Thompson and Sharp's Test: Weighted Linear Regression Random Effects Test} \label{Thompson}
A method proposed in \citet{thompson.sharp} allows for between study variance $\tau^2$, as introduced before in section \ref{sec:meta.analysis}. It extends the previously seen linear regression approach with $x = 1/s$ and $y = \theta/s$ by introducing weights. The effect size $\theta_i$ is assumed to be distributed as

\begin{align}
\theta_i \sim N(\beta_0 + \beta_1 s_i, s_i^2 + \tau^2) \label{t.sharp.regression}
\end{align}

$\tau^2$ is estimated as in equation \ref{eq:Tau.definition} (method of moments). %or \ref{eq:paule.mandel}. 
The weigths are set as $w_i = 1/\sqrt{s_i^2 + \tau^2})$. After adjusting for the weights as described in \ref{weighted.regression}, we can proceed analogous to Egger's test. The p-value for $\beta_{0} \neq 0$ reflects the evidence for a small study effect.

@@

The issue with dichotomous outcomes is that effect size and variance of effect size are correlated, which can readily be seen in \ref{eq:risk.ratio} and \ref{eq:risk.ratio.variance}. For example, a small number of event counts in one or group will inflate the variance and the effect size. Consequently, the tests above will tend to reject the null-hypothesis too often, i.e. report false positives. A number of solutions to this problem are provided in the literature.

\subsubsection{Peters Test: Weighted Linear Regression Test}
Instead of taking the standard error $s$ as explanatory variable $x$ as in Egger's Test, the inverse of the total sample size is used. Additionally, the variances $s_i^2$ are used as weights. Thus, the subsequent test procedure is identical to Egger's test. Peters test is a a small modification of Macaskill's test where the explanatory variable is the sample size instead of its inverse.

\subsubsection{Harbord's Test: Score based Test}
A rank based alternative to Peters test for binary outcomes is Harbord's test \citep{Harbord}.
It uses a different treatment effect and variance estimate: the score $\varphi$ of the log-likelihood, evaluated at log odds ratio $\theta_0 = 0$ and its inverse Fisher information $s^2$. Formally,

\begin{align}
\varphi &= e_t - (e_t - e_c)(e_t + (n_t - e_t))/(n_t + n_c) \label{harbord.score} \\
s^2 &= \frac{(e_t + e_c)(e_t + (n_t - e_t))(e_c + (n_c - e_c))((n_t - e_t) + (n_c - e_c))}{(n_t + n_c)^2(n_t + n_c - 1)} \label{harbord.variance}
\end{align}

It can be shown that they are both good approximations of the log odds ratio and its variance if the real $\theta$ is not too far from zero. The standardized estimator $r_i/v_i$ is also known as peto odds ratio. The obtained scores and variances can be used in Egger's test as treatment effects and variances.
